Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-02-21 ms.tex
fixed kinetic feedback section2

\begin{figure}
  \includegraphics[width=\columnwidth]{profiles.eps}
  \caption{\label{fig:profiles}Radial profiles of the CNM density
    ({\it top}), temperature ({\it middle}), and velocity ({\it bottom}),
    calculated for a representative sample of galaxies.  Different colors
    denote different values of the effective wind heating rate,
    $\vwO=1200$ km s$^{-1}$ ({\it blue}), $\vwO$=600 km s$^{-1}$ ({\it
      orange}), and 300 km s$^{-1}$ ({\it green}).  Different
    linestyles correspond to different values of $\Mbh$, $\Mbh=10^6
    \Msun$ ({\it dot-dashed}), $\Mbh=10^7 \Msun$ ({\it solid}), and
    $\Mbh=10^8 \Msun$ ({\it dashed}). Thin lines correspond to cusp
    galaxies ($\Gamma$=0.8) and thick lines correspond to core
    galaxies ($\Gamma$=0.1). Squares mark the location of
    the stagnation radius for each solution {\bf BDM: added colored labels for the $v_w$ value, e.g. a blue 1200 km/s directly to the top figure} .
 }
\end{figure}

@@

In Appendix \ref{app:windheat} we calculate the wind heating rate from stellar winds, $v_{\rm w}^{\star}$, and the mass loss parameter, $\eta$ (eq.~[\ref{eq:q}]), as a function of age, $\tau_{\star}$, of a stellar population which is formed impulsively (Fig.~\ref{fig:vwimp}).  At the earliest times ($\tau_{\star} \lesssim 10^{7}$ yr), the wind heating rate exceeds 1000 km s$^{-1}$, while much later ($\tau_{\star} \sim t_{\rm h}$) stellar wind heating dominated by main sequence winds is much lower, $v^{\star}_{\rm w} \sim 50-100 $ km s$^{-1}$.  We show in $\S\ref{sec:combined}$ that for the case of quasi-continueous star formation representative of the average star formation history of loss mass galaxies, the stellar wind heating rate produced by stellar winds can also be significant, $v_{\rm w}^{\star} \gtrsim 1000$ km s$^{-1}$.  Stellar wind heating thus contributes a potentially important source of both energy and mass for the CNM.

@@

The importance of Compton heating can be estimated by assuming the SMBH
radiates with a luminosity $L = \epsilon_{\rm rad} \dot{M}c^{2}$, where
$\epsilon_{\rm rad}$ is the radiative efficiency and where $\dot{M}$ is
estimated from equation (\ref{eq:mdot_analytic}).  Then using
equations (\ref{eq:rs_simple}), (\ref{eq:mdot_analytic}),
(\ref{eq:rhors}), (\ref{eq:rhostarrs}) we calculate that from equation
(\ref{eq:vw_eff}) that the effective Compton heating rate at the stagnation radius is given by
\begin{align} v_{w}^{\rm C} \simeq
  \begin{cases} 29 \eta_{0.02}^{0.5}T_{\rm
C,9}^{0.5}\epsilon_{-2}^{0.5} M_{\bullet,8}^{0.38}v_{500}^{-1.4}\,{\rm
km\,s^{-1}} &, \text{core}\\ 50 \eta_{0.02}^{0.5}T_{\rm
C,9}^{0.5}\epsilon_{-2}^{0.5} M_{\bullet,8}^{0.24}v_{500}^{-0.7}\,{\rm
km\,s^{-1}} &, \text{cusp},
  \end{cases}
  \label{eq:vC}
\end{align} where $T_{C,9} = T_{C}/10^{9}$ K and $\epsilon_{-2} =
\epsilon_{\rm rad}/0.01 \sim 1$ is the maximum radiative efficiency
for an advective accretion flow expected to produce high $T_{\rm C}$
emission (e.g.~\citealt{Narayan&Yi95}).  We caveat that, unlike stellar wind heating, Compton heating depends on radius as $v_{w}^{\rm C}(r) \propto (n^{2}\xi/\rho_{\star})^{1/2} \propto r^{(\Gamma-n-1)/2}$, i.e. $\propto r^{-0.6}$ and $\propto r^{-0.7}$ for core ($\Gamma = 0.8$; n = 1) and cusp ($\Gamma = 0.1$; n = 0.5) galaxies, respectively.  Thus, although the assumption of our numerical calculations that heating is radially constant is not satisfied, this radial dependence is sufficiently weak that we do not expect its presence to significantly alter our results.

If heating is provided by Compton heating alone, i.e. $\tilde{v}_{w} = v_{w}^{C}$, then equation
(\ref{eq:vC}) can be solved for $\tilde{v}_{\rm w}$ to find
\begin{align} v_{w}^{\rm C} \simeq
  \begin{cases} 150 \eta_{0.02}^{0.21}T_{\rm
C,9}^{0.21}\epsilon_{-2}^{0.21} M_{\bullet,8}^{0.16}\,{\rm km\,s^{-1}}
&, \text{core}\\ 130 \eta_{0.02}^{0.29}T_{\rm
C,9}^{0.29}\epsilon_{-2}^{0.29} M_{\bullet,8}^{0.14}\,{\rm km\,s^{-1}}
&, \text{cusp},
  \end{cases}
  \label{eq:vC2}
\end{align} 
The fact that $v_{w}^{\rm C} \gtrsim 300$ km s$^{-1}$ for physically realized values of $\eta \gtrsim 1$ shows that Compton heating can be significant in relatively young stellar populations with relatively high mass loss rates.

@@

Kinetic feedback results from outflows of energy or momentum in the form of a disk wind or jet from close to the black hole, which deposits its energy over much larger radial scales (e.g.~\citealt{McNamara&Nulsen07}; \citealt{Novak+11}; \citealt{Gaspari+12}).

Assume that the outflow power is proportional to the SMBH accretion rate, $L_{\rm j} =
\epsilon_{\rm j} \dot{M}c^{2}$, where $\epsilon_{\rm j} < 1$ is a jet efficiency factor, and that this energy is deposited as heat uniformly interior to a radius $r_{\rm heat}$ and volume $V_{\rm heat} \propto r_{\rm
heat}^{3}$.  The resulting volumetric heating rate $\dot{e} = \epsilon_{\rm
j}\dot{M}c^{2}/V_{\rm heat}$ near the stagnation radius produces an effective wind velocity given by
\begin{eqnarray} v_{w}^{\bullet} &\approx& \left(\frac{2t_{\rm
h}\epsilon_{\rm j}\dot{M}c^{2}}{V_{\rm heat}\eta
\rho_{\star}}\right)^{1/2} \sim \left(\frac{\epsilon_{\rm
j}\dot{M}c^{2}t_{\rm h}}{\eta M_{\rm enc}}\right)^{1/2} \nonumber \\
&\approx& 300\,{\rm km\,s^{-1}}\,\left(\frac{\epsilon_{\rm j}}{10^{-6}}\right)^{1/2}\left(\frac{r_{\rm s}}{r_{\rm
heat}}\right)^{1-\Gamma/2},
\label{eq:vjet}
\end{eqnarray} 
where we have used the facts that $\dot{M} = \eta M_{\star}|_{r_{\rm s}}/t_{\rm h}$ and, for radii inside the Nuker break radius, $M_{\star}(r) \propto
M_{\bullet}(r/r_{\rm soi})^{2-\Gamma}$ (eq.~[\ref{eq:rhostar}]).  If most of the outflow energy is released close to the stagnation ardius, $r_{\rm heat} \sim r_{\rm s}$, then even a small heating efficiency $\epsilon_{\rm j} \gtrsim
10^{-5}$ is sufficient for $v_{w}^{\bullet}$ to exceed other sources
of non-accretion powered heating.  However, if the disk outflow instead distributes energy over much larger scales comparable to the size of the galaxy, i.e. $r_{\rm heat} \gtrsim $ 10 kpc $\sim 10^{4}r_{\rm s}$, then even for an efficient jet, kinetic feedback is unimportant.

Indeed, using the results of \citet{Bromberg+11}, we estimate that a jet of luminosity $L_{\rm
  j}$ and half opening angle $\theta_{\rm j} = 0.1$ (characteristic of
AGN jets) will propagate through a gaseous mass $M_{\rm g}$ to a radius $r$ in a timescale
\begin{eqnarray}
t_{\rm jet} &\sim& 4000\,{\rm yr}\left(\frac{{L}_{\rm j}}{10^{40}\,\rm erg\,s^{-1}}\right)^{-1/3}\left(\frac{r}{\rm pc}\right)^{2/3}\left(\frac{M_{\rm g}}{10^{8}M_{\odot}}\right)^{1/3} 
\end{eqnarray}
If we write $M_{\rm g} \sim \dot{M}t_{\rm ff}$ then the ratio of the jet crossing time to the dynamical in the bulge $t_{\rm dyn} \sim r/\sigma$ is given by
\begin{equation}
\frac{t_{\rm jet}}{t_{\rm dyn}} \sim 7\times 10^{-3}M_{\bullet,8}^{0.13}\left(\frac{\epsilon_{\rm j}}{10^{-6}}\right)^{-1/3},
\end{equation}
independent of radius $r$.  Thus a jet with sufficient power, $\epsilon_{j} \gtrsim 10^{-6}$, to appreciably heat the CNM on radial scales $\sim r_{\rm s}$, also has sufficient power to escape the gas within the nucleus and propogate to much larger radii.  Slower outflows from the accretion disk, instead of a collimated relativistic jet, may provide a more promising source of feedback in these systems (e.g.~\citealt{Li+13}).

>>>>

\begin{figure}
  \includegraphics[width=\columnwidth]{profiles.eps}
  \caption{\label{fig:profiles}Radial profiles of the CNM density
    ({\it top}), temperature ({\it middle}), and velocity ({\it bottom}),
    calculated for a representative sample of galaxies.  Different colors
    denote different values of the effective wind heating rate,
    $\vwO=1200$ km s$^{-1}$ ({\it blue}), $\vwO$=600 km s$^{-1}$ ({\it
      orange}), and 300 km s$^{-1}$ ({\it green}).  Different
    linestyles correspond to different values of $\Mbh$, $\Mbh=10^6
    \Msun$ ({\it dot-dashed}), $\Mbh=10^7 \Msun$ ({\it solid}), and
    $\Mbh=10^8 \Msun$ ({\it dashed}). Thin lines correspond to cusp
    galaxies ($\Gamma$=0.8) and thick lines correspond to core
    galaxies ($\Gamma$=0.1). Squares mark the location of
    the stagnation radius for each solution {\bf BDM: added colored labels for the $v_w$ value, e.g. a blue 1200 km/s directly to the top figure.  what does v/cs look like?} .
 }
\end{figure}

@@

In Appendix \ref{app:windheat} we calculate the wind heating rate from stellar winds, $v_{\rm w}^{\star}$, and the mass loss parameter, $\eta$ (eq.~[\ref{eq:q}]), as a function of age, $\tau_{\star}$, of a stellar population which is formed impulsively (Fig.~\ref{fig:vwimp}).  At the earliest times ($\tau_{\star} \lesssim 10^{7}$ yr), the wind heating rate exceeds 1000 km s$^{-1}$, while much later ($\tau_{\star} \sim t_{\rm h}$) stellar wind heating dominated by main sequence winds is much lower, $v^{\star}_{\rm w} \sim 50-100 $ km s$^{-1}$.  We show in $\S\ref{sec:combined}$ that for the case of quasi-continueous star formation representative of the average star formation history of loss mass galaxies, the stellar wind heating rate produced by stellar winds can also be significant, $v_{\rm w}^{\star} \gtrsim 1000$ km s$^{-1}$.  Stellar winds thus contributes a potentially important source of both energy and mass for the CNM.

@@

The importance of Compton heating can be estimated by assuming the SMBH
radiates with a luminosity $L = \epsilon_{\rm rad} \dot{M}c^{2}$, where
$\epsilon_{\rm rad}$ is the radiative efficiency and where $\dot{M}$ is
estimated from equation (\ref{eq:mdot_analytic}).  Then using
equations (\ref{eq:rs_simple}), (\ref{eq:mdot_analytic}),
(\ref{eq:rhors}), (\ref{eq:rhostarrs}) we calculate that from equation
(\ref{eq:vw_eff}) that the effective heating rate at the stagnation radius is given by
\begin{align} v_{w}^{\rm C} \simeq
  \begin{cases} 29 \eta_{0.02}^{0.5}T_{\rm
C,9}^{0.5}\epsilon_{-2}^{0.5} M_{\bullet,8}^{0.38}v_{500}^{-1.4}\,{\rm
km\,s^{-1}} &, \text{core}\\ 50 \eta_{0.02}^{0.5}T_{\rm
C,9}^{0.5}\epsilon_{-2}^{0.5} M_{\bullet,8}^{0.24}v_{500}^{-0.7}\,{\rm
km\,s^{-1}} &, \text{cusp},
  \end{cases}
  \label{eq:vC}
\end{align} where $T_{C,9} = T_{C}/10^{9}$ K and $\epsilon_{-2} =
\epsilon_{\rm rad}/0.01 \sim 1$ is the maximum radiative efficiency
for an advective accretion flow expected to produce high $T_{\rm C}$
emission (e.g.~\citealt{Narayan&Yi95}).  We caveat that, unlike stellar wind heating, Compton heating depends on radius, scaling as $v_{w}^{\rm C}(r) \propto (n^{2}\xi/\rho_{\star})^{1/2} \propto r^{(\Gamma-n-1)/2}$, i.e. $\propto r^{-0.6}$ and $\propto r^{-0.7}$ for core ($\Gamma = 0.8$; n = 1) and cusp ($\Gamma = 0.1$; n = 0.5) galaxies, respectively.  The assumption of our model that the heating parameter be constant with radius is not satisfied, but this radial variation is sufficiently weak that it should not significantly alter our conclusions.

If Compton heating acts alone, i.e. $\tilde{v}_{w} = v_{w}^{C}$, then solving equation
(\ref{eq:vC}) for $\tilde{v}_{\rm w}$ yields
\begin{align} v_{w}^{\rm C} \simeq
  \begin{cases} 150 \eta_{0.02}^{0.21}T_{\rm
C,9}^{0.21}\epsilon_{-2}^{0.21} M_{\bullet,8}^{0.16}\,{\rm km\,s^{-1}}
&, \text{core}\\ 130 \eta_{0.02}^{0.29}T_{\rm
C,9}^{0.29}\epsilon_{-2}^{0.29} M_{\bullet,8}^{0.14}\,{\rm km\,s^{-1}}
&, \text{cusp},
  \end{cases}
  \label{eq:vC2}
\end{align} 
Compton heating can thus be significant in young stellar populations with relatively high mass loss rates, e.g. $v_{w}^{\rm C} \gtrsim 300$ km s$^{-1}$ for $\eta \gtrsim 1$.

@@

Kinetic feedback results from outflows of energy or momentum from close to the black hole in the form of a disk wind or jet, which deposits its energy as heat, e.g. via shocks or wave dissipation, over much larger radial scales (e.g.~\citealt{McNamara&Nulsen07}; \citealt{Novak+11}; \citealt{Gaspari+12}).

Assume that the outflow power is proportional to the SMBH accretion rate, $L_{\rm j} =
\epsilon_{\rm j} \dot{M}c^{2}$, where $\epsilon_{\rm j} < 1$ is a jet efficiency factor.  Further assume that this energy is deposited as heat uniformly interior to a radius $r_{\rm heat}$ and volume $V_{\rm heat} \propto r_{\rm
heat}^{3}$.  The resulting volumetric heating rate $\dot{e} = \epsilon_{\rm
j}\dot{M}c^{2}/V_{\rm heat}$ near the stagnation radius results in a heating parameter given by
\begin{eqnarray} v_{w}^{\bullet} &\approx& \left(\frac{2t_{\rm
h}\epsilon_{\rm j}\dot{M}c^{2}}{V_{\rm heat}\eta
\rho_{\star}}\right)^{1/2} \sim \left(\frac{\epsilon_{\rm
j}\dot{M}c^{2}t_{\rm h}}{\eta M_{\rm enc}}\right)^{1/2} \nonumber \\
&\approx& 300\,{\rm km\,s^{-1}}\,\left(\frac{\epsilon_{\rm j}}{10^{-6}}\right)^{1/2}\left(\frac{r_{\rm s}}{r_{\rm
heat}}\right)^{1-\Gamma/2},
\label{eq:vjet}
\end{eqnarray} 
where we have used the facts that $\dot{M} = \eta M_{\star}|_{r_{\rm s}}/t_{\rm h}$ and, for radii inside the Nuker break radius, $M_{\star}(r) \propto
M_{\bullet}(r/r_{\rm soi})^{2-\Gamma}$ (eq.~[\ref{eq:rhostar}]).  If the bulk of the outflow energy is released near the stagnation radius, then even a small heating efficiency $\epsilon_{\rm j} \gtrsim
10^{-5}$ is sufficient for $v_{w}^{\bullet}$ to exceed other sources of non-accretion powered heating.  However, if this energy is deposited over much larger physical scales comparable to the size of the galaxy, i.e. $r_{\rm heat} \gtrsim $ 10 kpc $\sim 10^{4}r_{\rm s}$, then kinetic feedback is unimportant, even for a powerful jet with $\epsilon_{\rm j} \sim 1$.

Using the results of \citet{Bromberg+11}, the time required for a jet of luminosity $L_{\rm
  j}$ and half opening angle $\theta_{\rm j} = 0.1$ (characteristic of
AGN jets) to propagate through a gaseous mass $M_{\rm g}$ of radius $r$ is estimated to be
\begin{eqnarray}
t_{\rm jet} &\sim& 4000\,{\rm yr}\left(\frac{{L}_{\rm j}}{10^{40}\,\rm erg\,s^{-1}}\right)^{-1/3}\left(\frac{r}{\rm pc}\right)^{2/3}\left(\frac{M_{\rm g}}{10^{8}M_{\odot}}\right)^{1/3} 
\end{eqnarray}
Approximating $M_{\rm g} \sim \dot{M}t_{\rm ff}$, the ratio of the jet escape timescale to the dynamical timescale $t_{\rm dyn} \sim r/\sigma$ is given by
\be
\frac{t_{\rm jet}}{t_{\rm dyn}} \sim 7\times 10^{-3}M_{\bullet,8}^{0.13}\left(\frac{\epsilon_{\rm j}}{10^{-6}}\right)^{-1/3},
\ee
independent of $r$.  A jet with power sufficient to appreciably heat the CNM on radial scales $\sim r_{\rm s}$ ($\epsilon_{j} \gtrsim 10^{-6}$) also necessarily has sufficient power to escape the nucler region and propagate to much larger radii.  Slower outflows from the accretion disk, instead of a collimated relativistic jet, may provide a more promising source of feedback in these systems (e.g.~\citealt{Li+13}).
